---
title: "SWOTlists"
author: "Mark Hagemann"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## DAWG matrices

SWOT-like datasets contain variables defined at specific points in time and space. The Discharge Algorithm Working Group (DAWG) has adopted the convention of formatting such data as space-down, time-across matrices. I will refer to such a space-down, time-across matrix as a DAWG matrix. `swotr` contains a handy function for plotting DAWG matrices; that function is `plot_DAWG`. For example, we can generate a random DAWG matrix indexed in 4 locations and 5 times as follows:

```{r}
library(dplyr)
library(ggplot2)
library(swotr)
mat1 <- matrix(rnorm(20), nrow = 4, ncol = 5)

plot_DAWG(mat1)
```

Variables that vary only in time or only in space can be converted from vectors to DAWG matrices using the `swot_vec2mat` function. For example:

```{r, fig.show = 'hold'}
time_vec <- rnorm(5) # vector of a temporal variable
space_vec <- rnorm(4) # vector of a spatial variable

time_mat <- swot_vec2mat(time_vec, pattern = mat1) # mat1 is used as a pattern to get the matrix dimensions
space_mat <- swot_vec2mat(space_vec, pattern = mat1) # automatically determines which dimension is variable

plot_DAWG(time_mat) + ggtitle("temporal variable")
plot_DAWG(space_mat) + ggtitle("spatial variable")
```

Unfortunately, `swot_vec2mat` is too dumb to work if `pattern` has the same number of rows as columns. 

## swotlists

A swotlist is essentially a named list of DAWG matrices. This is not a formal object class and is flexible as to what variables it contains. It's just a kind of list that many `swotr` functions take as input. For example, the following is a perfectly valid swotlist. 

```{r}
list1 <- list(var1 = mat1, var2 = mat1 + rnorm(20))
```

The `swot_plot` function plots all variables (or optionally a subset thereof) within a swotlist. 

```{r}
swot_plot(list1)
```

A swotlist is a compact, flexible, and easily accessible way of storing SWOT-like data. But it is limited in that it can only contain matrices with the same dimensions as one another. Thanks to easy matrix representation of vectors (via `swot_vec2mat`) swotlists can represent a surprising variety of variables. 

Although there's no requirement to what a swotlist can contain, some `swotr` functions (and other functions that may be added to swotr or released in a companion package in the future) expect certain DAWG matrices to be present. 

## Tidying and untidying swotlists

Although a handy data format, swotlists are not perfect for using with many R utilities that expect `data.frame`s and related "tidy" tabular data representations. (See [the tidyverse](https://www.tidyverse.org/) for some examples of how tidiness has permeated the R ecosystem.) `swotr` has two functions for switching between swotlist representation and tidy representation of SWOT-like data. To turn a swotlist into a tidy `data.frame`, use `swot_tidy`.

```{r}
tidydf1 <- swot_tidy(list1)
str(tidydf1)
```

To go back to the original swotlist for, use `swot_untidy`. 

```{r}
str(swot_untidy(tidydf1))
```

`swot_tidy` is used under the hood for many of `swotr`'s plotting functions (and some other functions), since ggplot2 takes a `data.frame` as input.

## netcdfs

`swotr` contains functions to make reading SWOT-like netcdf files easy. But in order to use these you'll need to have the `ncdf4` package installed. This in turn requires that you have the netcdf library installed on your computer. 

The `nc_list` function reads all data from a netcdf file into a single list. 

```{r}
ncfile <- system.file("extdata", "SacramentoDownstream.nc", package = "swotr")

saclist1 <- nc_list(ncfile)

str(saclist1)
```

Note that this is not a swotlist, as it contains elements that are not matrices, and it contains matrices that have differing numbers of rows (9 for reach data, 93 for cross-section data). But it contains all of the data in the netcdf should you need it.

Since the McFLI community works principally with reach-averaged data, a useful alternative to `nc_list`, `nc_reach`, is proviced. Unlike `nc_list`, `nc_reach` returns a swotlist of only reach-averaged quantities. 

```{r}
saclist2 <- nc_reach(ncfile)
str(saclist2)
```

Note that `saclist2` also includes a `QWBM` attribute. This is to store the water balance model (WBM) estimate of discharge. Since this is a scalar, it would be extremely redundant to store as a DAWG matrix, but it is often useful to store with the other reach variables. Keeping it as an attribute means the object structure is not affected, and it is still a swotlist containing only DAWG matrices. 

Since `saclist2` is a swotlist, we can plot it as follows:

```{r}
swot_plot(saclist2)
```

To get a better detail of area, width, and slope, we can plot just those variables on a log axis as follows

```{r}
swot_plot(saclist2, vars = c("A", "W", "S")) + scale_y_log10()
```


An optional argument, `good_only = TRUE` can be passed to `nc_reach`; this will read in only reaches flagged as "good" in the netcdf file. Since all reaches are "good" in the Downstream Sacramento, this does not affect this particular case. 

```{r}
saclist2 <- nc_reach(ncfile, good_only = TRUE)
```

## Subsetting swotlists

In the above plot it appears there are 4 potentially anomalous slope values in reach 1. It would be helpful to look just at that location's data to find out more. We can use the `swot_sset` function to subset an entire swotlist by time and/or location, and then run `swot_plot` on the subsetted result.

```{r}
saclist2 %>% 
  swot_sset(keeplocs = 1) %>% 
  swot_plot(vars = c("A", "W", "S"))
```

Indeed, we may want to remove those anomalous times. There are a couple of ways to do this. First, using `swot_sset`:

```{r}
badtimes <- which(saclist2$S > 6e-4, arr.ind = TRUE)[, 2] # Get just the columns (times) with S > 6e-4
saclist3 <- swot_sset(saclist2, keeptimes = -badtimes) # Use minus sign to omit, rather than keep, badtimes

swot_plot(saclist3)
```

And now the anomalous slopes are gone, as well as all other variables' data from those 4 times. 

If we thought the entire reach was bad we can removed 1 of 9 reaches instead of 4 of 154 times:

```{r}
saclist4 <- swot_sset(saclist2, keeplocs = -1) # again, negative to keep everything except 1

swot_plot(saclist4)
```

Now there are only 8 reaches, and the remaining reaches have been renumbered to replace the former reach 1. 

Another way to do the same operations is to replace the anomalous values with `NA`s and use `swot_purge_nas` on the result.

```{r}
saclist5 <- saclist2 # Make a copy before replacing values
saclist5$S[saclist5$S > 6e-4] <- NA
```

It may be perfectly fine to just keep the dataset with missing (`NA`) values, but some functions that expect complete data will not work. Purging of times or locations with `NA`s present is easy with `swot_purge_nas`:

```{r}
saclist6 <- swot_purge_nas(saclist5, purge = "times") # Result is equivalent to saclist3
saclist7 <- swot_purge_nas(saclist5, purge = "locs") # Result is equivalent to saclist4
```

## dA operations

`swotr` contains a few functions for working with partial cross-sectional area (`dA`) observations. 

First, `dA` can be calculated from height and width observations via `calcdA_vec` (for single-location vectors) and `calcdA_mat` (for DAWG matrices). ***Note that the summation-as-integration method used by these functions is not robust to height and width measurement error and should be used with this imprecision in mind.***

```{r}
dAmat <- calcdA_mat(w = saclist2$W, h = saclist2$H)
plot_DAWG(dAmat)
```

Since the zero reference of `dA` is arbitrary, `swotr`'s dA-related functions include an argument specifying where to set `dA = 0`. Options are "first" for the first observation in time  (this is typically the default), "minimum" or "median". Existing DAWG matrices of `dA` can have their zero point adjusted via `rezero_dA`. 

```{r, fig.show = "hold"}
plot_DAWG(rezero_dA(dAmat, zer = "minimum"))
plot_DAWG(rezero_dA(dAmat, zer = "median"))
```

The full cross-sectional area is obtained by adding `dA` to $A_0$, the constant-in-time unobserved part of cross-sectional area. The value of $A_0$ depends on where dA is referenced to zero. If a swotlist contains a cross-sectional area DAWG matrix called `A`, the true $A_0$ vector can be obtained using the function `realA0`. This also has a `zero` argument, which should match the zero reference of the swotlist's `dA` matrix. 

```{r}
realA0(saclist2, zero = "first")
```

## Interfacing with bamr

The `swotr` package is designed to interface with the [bamr](http://github.com/markwh/bamr) package for estimation of flow-law parameters and discharge. To turn a swotlist into a `bamdata` object, use swot_bamdata:

```{r}
sacbd <- swot_bamdata()
```

